Critical assessment of Linear-Response Excitations in Molecular system\\ or \\ On the Localization of the Linear-Response excitations in Open Systems

### 2018-01-17 LRDraft.tex
update

We split the unperturbed Hamiltonian as the kinetic $\op T$ plus the potential $\op V$ term, so that the equation of motion for $\ket{f_p(\omega)}$ is recasted as:
\be\lb{fluctuationStateEqMotion}
\left(\bbomega+\eps_p -\op T\right)\ket{f_p(\omega)} = \op V\ket{f_p(\omega)} + \op Q_0\left(\op V'(\omega) + \op \Phi(\omega)\right)\ket{\psi_p} \;.
\ee
This equation has the formal structure of an inhomogeneous Helmholtz equation and can be written in the coordinate representation as:
\be
\left(k_p(\omega)^2+\ii\eta + \nabla^2\right)f_p(\omega,\r) = s(\omega,\r) \;\;, {\rm with} \;\; k_p(\omega)^2 = 2(\omega+\eps_p) \;,
\ee
and admits an integral representation in the form
\be\lb{fluctuationStateIntEq1}
f_p(\omega,\r) = \int \dd\r' g_H(\r-\r')s(\r') \;,
\ee
where $g_H(\r)$ is the Green function of the Helmholtz operator in three dimensions. This integral representation allows us to assess the localization character 
of the fluctuation states, \emph{i.e.} the behavior of $f_p(\omega,\r)$ when $\r$ becomes large. Indeed, assuming that the support of the source term is compact
we have that, for $\r$ outside the domain in which the integrand of \eqref{fluctuationStateIntEq1} is not vanishing:
\be\lb{fluctuationStateIntEq2}
f_p(\omega,\r) \sim \frac{e^{\ii \sqrt{k_p(\omega)^2} |\r|}}{|\r|} \;,
\ee
so the asymptotic character of the fluctuation states depends on the sign of $k_p(\omega)^2$. In particular equation \eqref{fluctuationStateIntEq2} describes an
exponential damping for $\omega < |\eps_p|$ and an oscillating power law behavior for $\omega \geq |\eps_p|$ ().

@@

% We present an ulterior expression of the response density operator in the frequency space with the aim to establish its (spatial) localization properties.
% The starting point of this analysis is the expression of the response density in terms of the commutator of \eqref{rhoPrimeOmegaDef1}. We observe that, for any 
% operator $\op O$, we may write:  
% \be
% \commutator{\dmnot}{\op O} = \dmnot \op O \op Q_0 - \op Q_0 \op O \dmnot\;,
% \ee
% where $\op Q_0= \identity - \dmnot$ projects the operator in the subspace orthogonal to $\dmnot$. So, thanks to this relation we have that:
% \be
% \dm'(\omega) = \ii\sum_p\left(\ketbra{\psi_p}{\psi_p}\op A(\omega)\op Q_0 - \op Q_0\op A(\omega)\ketbra{\psi_p}{\psi_p}\right) \;.
% \ee
% This equation suggests us to define the \emph{fluctuation states}:
% \be%\lb{fluctuationStateDef1}
% \ket{f_p(\omega)} \doteq -\ii\op Q_0\op A(\omega)\ket{\psi_p} \;, \qq
% \bra{f_p(\omega)} \doteq \bra{\psi_p}\ii\op A(-\omega)\op Q_0 \;,
% \ee
% where the $\omega$ dependence of the bra is consistent with hermitian conjugation property of the Fourier transform of a hermitian operator. Fluctuation states are, by definition,
% orthogonal to the occupied eigenstates of $\hnot$, so that:
% \be
% \brket{f_p(\omega)}{\psi_q} = \brket{\psi_q}{f_p(\omega)} = 0 \;, \forall \, p,q,\omega
% \ee
% The response density written in terms of the fluctuation states reads:
% \be\lb{rhoPrimeFluctuationStateDef1}
% \dm'(\omega) = \sum_p\left(\ketbra{\psi_p}{f_p(-\omega)} + \ketbra{f_p(\omega)}{\psi_p}\right) \;.
% \ee

>>>>

We split the unperturbed Hamiltonian as the kinetic $\op T$ plus the potential $\op V$ term, so that the equation of motion for $\ket{f_p(\omega)}$ is recasted as:
\be\lb{fluctuationStateEqMotion}
\left(\bbomega+\eps_p -\op T\right)\ket{f_p(\omega)} = \op V\ket{f_p(\omega)} + \op Q_0\left(\op V'(\omega) + \op \Phi(\omega)\right)\ket{\psi_p} \;.
\ee
This equation has the formal structure of an inhomogeneous Helmholtz equation and can be written in the coordinate representation as:
\be
\left(k_p(\omega)^2+\ii\eta + \nabla^2\right)f_p(\omega,\r) = s(\omega,\r) \;\;, {\rm with} \;\; k_p(\omega)^2 = 2(\omega+\eps_p) \;,
\ee
and admits an integral representation in the form
\be\lb{fluctuationStateIntEq1}
f_p(\omega,\r) = \int \dd\r' g_H(\r-\r')s(\r') \;,
\ee
where $g_H(\r)$ is the Green function of the Helmholtz operator in three dimensions. This integral representation allows us to assess the localization character 
of the fluctuation states, \emph{i.e.} the behavior of $f_p(\omega,\r)$ when $\r$ becomes large. Indeed, assuming that the support of the source term is compact
we have that, for $\r$ outside the domain in which the integrand of \eqref{fluctuationStateIntEq1} is not vanishing:
\be\lb{fluctuationStateIntEq2}
f_p(\omega,\r) \sim \frac{e^{\ii \sqrt{k_p(\omega)^2} |\r|}}{|\r|} \;,
\ee
so the asymptotic character of the fluctuation states depends on the sign of $k_p(\omega)^2$. In particular equation \eqref{fluctuationStateIntEq2} describes an
exponential damping for $\omega < |\eps_p|$ and an oscillating power law behavior for $\omega \geq |\eps_p|$.

@@

\subsection{Criterion for the choice of the underlying computational basis}

Let us further investigate formula \eqref{LinearResponseFunctDef2} pointing out some computation aspects. We introduce a computational basis $\{\ket{\mu}\}$, composed by a \emph{finite}
set of functions on which we expand the formal scalar products appearing in \eqref{LinearResponseFunctDef2} as follows:
\be
\bra{f_p(-\omega)}\op O\ket{\psi_p} = \sum_\mu \brket{f_p(-\omega)}{\mu}\bra{\mu}\op O\ket{\psi_p} \;.
\ee
The quality of the basis can be evaluated by the requirement that the numerical coefficients appearing in the r.h.s provide a reliable reconstruction of the formal scalar product. 
Here we are assuming that the ground state properties of the unperturbed system have been computed in the same basis introduced above. This means that the feature of the basis, 
that for instance for real space basis are given by the grid spacing and number of points (that determines the dimension simulation box), are chosen so that the coefficients
$\brket{\mu}{\psi_p}$ represent a valid discretization of the occupied orbitals. The same can be stated also for the matrix elements $\bra{\mu}\op O\ket{\psi_p}$, by assuming 
that $\op O$ acts as a local operator on the occupied orbitals.

The most interesting point of this discussion concerns the assessment of the reliable sampling of the fluctuation states through the coefficients $\brket{f_p(-\omega)}{\mu}$ and
$\brket{\mu}{f_p(\omega)}$. The question is non-trivial, as the behavior of the fluctuation states strongly depends on $\omega$, as previously explained.
The first step needed to address this problem concerns with the evaluation of the reliability of the $\{\ket{\mu}\}$ basis for $\omega=0$, \emph{i.e.} for a perturbing field that does 
not depend on time. In this case the fluctuation states, as well as the expectation value of a given observable, can be computed performing a ground state calculation in which the perturbing 
field $\op\Phi$ is added to the external potential of $\hnot$\footnote{We are interested in the linear response, so the magnitude of $\op\Phi$ has to be chosen in a way that respects this 
regime. To check this point we can compute the perturbation-induced modification of an observable an verify that its value scales linearly with the magnitude of the field}. %, without

%In order to address this aspects it is possible to assess the reali
%We propose an operative procedure to perform a direct measurement of the

%An explicit check can be performed at $\omega=0$.

%This condition implies that the coefficients
%$\brket{\mu}{\psi_p}$ represent a valid discretization of the occupied orbitals and also the matrix elements $\bra{\mu}\op O\ket{\psi_p}$ can assumed to be well sampled, assuming that
%$\op O$ acts as a local operator on the occupied orbitals.
